Differential chemotaxis with competition as a mapping mechanism in the brain

### 2022-09-27 paper/main.tex
Changes to paper which is now almost in submittable form

\emph{The retinotectal projection, in which the axons of retinal ganglion cells form an ordered map in the optic tectum, is an important developmental system that has been studied for many decades. There is a large body of evidence, consisting of surgical and genetic manipulations, that give insight into how it behaves. Experimental work has uncovered a number of mechanisms that are involved in aspects of the mapping. However, a single, consistent theoretical framework has yet to describe all the patterns that are formed. Sperry's chemoaffinity theory provides an account of the wildtype pattern. Gierer formalised and extended this theory to incorporate competition to account for a subset of surgical manipulation experiments, but it was never tested against the full range of surgical manipulations. Other theoretical and computational studies have been partially successful in accounting for surgical manipulations, but to date, no single theory, operating only by local interactions, has accounted for every manipulation. We present a two-dimensional model of the retinotectal projection resembling that of Gierer, in which orthogonal pairs of opposing tectal ligand gradients combined with retinal gradients of receptors form Gierer's potential functions and show that this receptor-ligand model accounts for the full range of surgical manipulations. We then extend the detail with which the receptor-ligand interactions are treated to account for more recent genetic manipulation experiments.  }

%\begin{itemize}
%    \item Sperry's chemospecificity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development.
%    \item A key component of the theory is that growing axons find their way into position by chemotaxis, i.e., by evaluating local information in gradients of gene expression that span the target tissue
%    \item Computational models have shown that in addition to chemotaxis, local competitive interactions between growing axons can account for a wide range of results from experiments in which the factors known to influence the growth of axons have been surgically or genetically manipulated
%    \item However, while such models have \emph{assumed} the effects of chemotaxis, the key assumption of the chemospecificity model, that axons sample and evaluate positional information only \emph{locally}, is yet to be tested directly.
%    \item Consider, for example, an important model by Goodhill \& Simpson, which shows how [a certain set of assumptions about the form of the] competition between axons can account for patterns of innervation in the developing tectum that result from ablating/rotating sections of the retina. 
%    \item The formulation of this model includes a term that at every time-step moves the simulated axons in the direction of a pre-specified target location. Thus the model implicitly assumes a global supervisor, capable of evaluating the current state of each axon, computing a direction vector to correct its trajectory, and communicating that information to each axon.
%    \item This model was constructed primarily to establish the effects of competition, and to this end is elegantly formulated.
%    \item Nevertheless, the key assumptions of the chemospecificity model, and specifically the proposal that axons find their way via evaluating only local information (chemotaxis), are yet to be investigated directly.
%    \item The aim of the current study is therefore to construct a model that represents the assumptions of the chemospecificity model explicitly. 
%    \item One advantage of such a model is that it can be used to investigate the sensitivity and robustness of local interactions in establishing topographic maps to various sources of noise.
%    \item Indeed, our simulation results show that the local mechanisms of axonal information-processing that are assumed by the chemospecificity model provide a remarkably robust solution to the problem of topographic map development.
%\end{itemize}

@@

Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity (\textbf{find citation}).
In this system a two-dimensional map of the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum. 
This retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum).
This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

@@

Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}. 
%\emph{PDEs too complex (can't do maths analysis), so need agent based. Can represent more biological detail and ask how that affects the systems level.}
These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue. 
In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples). 
However as the number of terms involved in PDEs necessarily grows large in systems representing differential chemotaxis, their analytical tractability quickly disappears, making the potential advantage of such modelling redundant.
Additionally, the numerical simulation of a PDE system can become intractable, because each axonal projection must be represented by several state variables, each of which require solution across the entire spatial domain (which may consist of many thousands of elements).

An alternative simulation approach is to treat each of $n$ axons as an agent that interacts with its environment and with the other $n-1$ axons. 
The computation of such a system (see for example, \citet{simpson_simple_2011}) is dominated by the $O(n^2)$ axon-axon interactions and so a limit will be approached as $n\rightarrow \infty$. 
We found that agent based simulations of $O(10^3)$ interacting axons can be achieved on a standard desktop computer whereas an equivalent PDE system would fail with about 100 axon projections.
A further advantage of an agent based simulation is that the addition of new biological details is easy to implement and comes an a minimal computational cost, whereas in a PDE system, each new detail requires at least one state variable, which must be solved across the entire domain.

% List questions that emerge from the above
For these reasons, we sought to implement an agent-based differential chemotaxis model of the retinotectal projection to ask: i) Can a two-dimensional implementation of Gierer's model explain the full range of surgical manipulations that have been described in the literature? ii) Can the model be developed such that only \emph{local} interactions are required to transfer information between agents? and iii) Can a differential chemotaxis model explain more recent results in which levels of EphA receptor sub-types are changed by genetic manipulation \citep{brown_topographic_2000,reber_relative_2004}?

@@

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (Supplementary Fig.\,\ref{f:experimental_expressions} summarises results from the literature). 
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}).

@@

After finding an optimum set of parameters for the wildtype arrangement, we tested our model with a number of simulations of surgical manipulations (for reviews, see \citet{udin_formation_1988} and \citet{goodhill_retinotectal_1999}). 
Figs.~\ref{f:GJsurg_grafts}A and \ref{f:GJsurg_grafts}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}. 
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column). 
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression. 
The manipulated gradients, interacting with unchanged axon branches, lead to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours. 
Long connecting lines in the fish net show that a number of axons from near the edges of the tectum have been drawn into the graft region. 
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg_grafts}A \& B. 
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red). 
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04). 
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg_grafts}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped. 
There is considerable disorder in the simulated pattern of axon centroids but a line of red/yellow agents that would been located caudally has advanced rostrally and a number of green agents are more caudal than in the wildtype arrangement, consistent with experimental observations.
% Note: the graph shows the pattern metric for a region that covers BOTH grafts and for the region that is not the two grafts.

Fig.\,\ref{f:GJsurg_ablate}A shows the result of a retinal ablation, in which the temporal half of the retina is surgically removed and the remaining RGC axons are allowed to develop on the unmanipulated tectum.
According to a pure chemoaffinity theory, the surviving retinal axons should retain their chemical labels and they would be expected to find their usual locations on the tectum, resulting in a tectum which is only half populated with axons. 
The experimental observation is that the axons arrange themselves across the whole tectum, retaining their topographic order (inset, \citet{attardi_preferential_1963,schmidt_expansion_1978}). 
The fish-net plot of the simulation indicates that the competition in the system successfully `stretches out' the surviving axons to achieve a close representation of the experimentally observed result. 
In this figure, the right-hand graph shows the simulated position of each axon centroid on the rostro-caudal tectal axis plotted against its position on the nasal-temporal retinal axis in red triangles, with the experimental positions plotted in blue squares.
%
Fig.\,\ref{f:GJsurg_ablate}B shows the result of a tectal ablation. Here the retina is left un-modified, but the caudal half of the tectum is surgically removed. Experimental observation \citep{yoon_reorganization_1971,sharma_reformation_1972} shows that RGC axons `crush up' on the remaining rostral section, retaining topographic order and creating a complete map on the available tectal tissue. Our model clearly replicates this result, indicated by both fish-net and tectal/retinal axes graphs, because the chemotaxis component of the model is based on balanced, relative interactions between opposing gradients, rather than absolute labels or expression measurements.
%
A related ablation experiment is the so-called `mismatch', in which the temporal half of the retina is ablated along with the caudal part of the tectum to which it normally maps, again challenging the pure chemoaffinity model \citep{horder_retention_1971}. Our model replicates this manipulation (Fig.\,\ref{f:GJsurg_ablate}C) for the same reason as in the simple tectal ablation of Fig.\,\ref{f:GJsurg_ablate}B; opposing gradients remain on the surviving tectum allowing RGC axons to find their relative ordering.
%
The final surgical manipulation which we investigated was the `compound eye' experiment, in which two nasal retina halves are fused together \citep{gaze_retino-tectal_1963,fawcett_retinotectal_1982}. Within the compound retina there exist matched pairs of retinal ganglion cells, one from each nasal half, whose receptor and ligand expressions cannot be distinguished. These form overlaid maps, each of which spans the entire tectum (Fig.\,\ref{f:GJsurg_ablate}D).

@@

In the early 2000s, a series of genetic experiments added to the set of manipulations that a model of the retinotectal system would need to
explain. 
First, \citet{brown_topographic_2000} showed that a selective RGC knock-in of EphA3 receptors led to an interesting modulation of the maps found in the tectum. 
For the strongest knock-in (a homozygote manipulation) two maps were formed, with knock-in cells finding termination zones more rostral than their wildtype retinal neighbours. 
By enhancing the quantity of EphA receptors expressed by half of all retino-tectal axons, they demonstrated the importance of retinal receptor expression for the arrangement of the tectal map. 
In the weaker, heterozygote manipulation, the maps were still separated, but the separation `collapsed' towards the rostral end of the tectum (see \citet{brown_topographic_2000}, Fig\,5). 
This led to a second study showing that with an additional knock-down of EphA4 receptors in \emph{all} RGCs, the retinal origin of the collapse point was shifted temporally and all heterozygote EphA3 knock-in cells were pushed even further rostrally (\citet{reber_relative_2004}, their Fig.\,3e). 
For our model to successfully reproduce these results, it should therefore: i)~Demonstrate a complete separation of maps for a `large' knock-in of EphA3; ii)~Show a separation of maps for a smaller knock-in of EphA3 that is greater for axons of nasal origin. iii)~For the smaller knock-in, a collapse point at 76\% of the N--T axis; iv)~Show a considerable rostral shift of cells for which EphA3 is knocked in and EphA4 is knocked down; and v)~Show a collapse point for the knock-in/knock-down cells and knock-down only cells at around 88\% of the N--T axis.

An assumption made in these papers is that all EphA receptor sub-types contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side. 
The precise contribution of each sub-type to the overall level is assumed not to be important. 
In our model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig\,\ref{f:ex}B) and so we treated $r_0$ as being equivalent to $\sum\mathrm{EphA}$ and manipulated it to attempt a recreation of the \citet{brown_topographic_2000}~and \citet{reber_relative_2004}~experiments. 
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 0.8 (heterozygote EphA3 ki/+) or 3.2 (homozygote EphA3 ki/ki). 
To include the additional heterozygote knock-down of EphA4 (EphA4 +/-) from \citet{reber_relative_2004}, we subtracted 1.2 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, a plot of axon centroids and final rostral-caudal position on the tectum versus the originating nasal-temporal retinal coordinate for time step 1500. 
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue. 
The right hand panel mapping final R--C position versus N--T origin  (and its inset which shows the experimental result) follow the same red/blue colour scheme.
The map of treated cells (red) was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because the curves remain separated for all N--T values. Regardless of the value of the ratio between treated and untreated cells, treated temporal axons are drawn to a significantly more rostral position than those without the knock-in. A bootstrapped studentized t-test of the two sets of points indicates a low probability, $p<10^{-4}$, that they are drawn from statistically identical distributions.
% While the convergence of the manipulated and wildtype lines was less abrupt than in the experimental result of \citet{brown_topographic_2000}, it was comparable with the result from \citet{simpson_simple_2011}.

Fig.\,\ref{f:GJeph}B shows the same trio of graphs for the homozygote double knock-in, EphA3 ki/ki. Here, the treated axons were pushed further towards the rostral end of the tectum and the untreated axons were pushed caudally due to competition with the treated cells. 
The maps were completely separated. As in the experiment, there was no collapse point.
The rostral movement of treated axons occurred due to their increased expression of EphA ($r_0$), which increased the simulated interaction with the tectal rostral-caudal ligand gradient. 
The opposing receptor expression ($r_2$) was unchanged by the manipulation and so those axons with increased $r_0$ expression were pulled to a more rostral position than they would otherwise have taken. 
The caudal movement of wildtype axons, whose receptor expression in Figs.\,\ref{f:GJeph}A-B was unchanged, occurred due to competition with the most caudal of the rostrally-drawn treated axons.
Overall, the correspondence between the experimental result and the simulated homozygous knock-in is close.

Figs.\,\ref{f:GJeph}C show the results for the \citet{reber_relative_2004} manipulation, in which the EphA expression was selectively enhanced (by EphA3 knock-in) in 50\% of cells and knocked down (by EphA4 knock-down) in all cells.
As in Fig.\,\ref{f:GJeph}A, the red and blue lines are separated, but there is no map collapse ($p<10^{-4}$). 
The significant rostral movement of the treated axons seen in \citet{reber_relative_2004}, where not a single treated axon was found more than 30\% along the rostral-caudal axis, was absent from this result. 
In fact, the knock-in/knock-down cells find locations slightly more \emph{caudal} than those in Fig.\,\ref{f:GJeph}A because while $r_0$ is increased by 0.8 from the knock-in, the knock-down of 1.2 reduces these cells' interaction with tectal ephrinA ($L_0$) compared with those in the EphA3 knock-in/EphA4 wildtype simulation.

It is important to note that these results emerge purely from modifications to the receptor expression levels; the model is otherwise unchanged. 
The lack of collapse points or the more rostral positions of knock-in/knock-down cells suggests that there is an interaction effect between EphA4 and other EphA receptors that has not been captured by this model.
We address possible interaction effects in the next section.

@@

$c_i\,r_{i,b}$ is the effective signal strength, $s_i$, for receptor $i$ and differs from $r_{i,b}$ only for $i=0$ for which (dropping the branch subscript) $s_0 = r_0/r_{\!\scriptscriptstyle A4}$.
Under this relationship, the signal strength changes as $r_0$ increases (simulating the knock-in of EphA3) and also as $r_{\!\scriptscriptstyle A4}$ decreases (simulating EphA4 knock-down).  
%(This is reminiscent of the divisive relationship suggested by the work of Brown, Reber and co-workers who showed that when the ratio of EphA in treated to EphA in untreated cells drops below a threshold, the mapping system directs the cells to the same tectal location.)
Fig.\,\ref{f:clustermech}C shows $s_0$ under knock-in, knock-down and combined knock-in/knock-down conditions.
The average magnitude of the signal $s_0$, as defined above, is smaller than $r_0$, which is the effective signal in the original model (in Eq.\,\ref{e:G}, $r_{i,b} \equiv s_i$). To counterbalance this reduction in the magnitude of the receptor-ligand interaction between $r_0$ and $l_0$, we reduced the value of $r_2$: $r_2 \rightarrow r_2/2$ which ensures that the cluster size model reproduces the wildtype result as in Fig.\,\ref{f:GJ}D.

@@

% Summarize the findings
Can the assumptions implicit in Sperry's chemoaffinity theory account for the behaviour of the axons that grow from the retina into the optic tectum?
If not, does the addition of competition resolve the remaining problems? 
Can chemotaxis and competition be implemented by invoking a signalling system that behaves like the known signalling molecules (Ephs/ephrins, Ryk/Wnt3 \& Neogenin/RGM)?
To answer these questions, we developed a model combining a gradient-following form of chemotaxis (Eq.\,\ref{e:G}) together with axon-axon competition (Eq.\,\ref{e:X}) in which information between axonal growth cone agents was only transferred by the local interactions between their receptors and ligands.
The model contained parameters to i) control the balance of competition and chemotaxis, ii) set the interaction range of the competition and iii) set a competitive interaction threshold. 
These parameters were optimized so that the model would reproduce the experimentally observed topographic patterning of growth cones seen in healthy animals. 
The model then correctly recreated a broad range of patterns that occur when healthy animals are subjected to surgical manipulations to disrupt neuronal development. 
Qualitative rotation was observed in graft rotation simulations; the patch swap simulation was similarly qualitatively accurate. 
Retinal and tectal ablation simulations produced highly convincing results, as did mismatch and compound eye simulations.

@@

% Place findings in context

@@

%\subsection*{Genetic manipulations}
% An assumption of the simulation of genetic manipulations was that there is no difference between the EphA receptor sub-types.

@@

% FIGURE S1
\begin{figure}
\includegraphics[width=\linewidth]{./images/Tissuevisb.png}
\caption{(Supplementary) Ligand expression and gradients on the tectum for the
graft-rotate 90 degrees simulated surgical manipulation}
\label{f:trot90}
\end{figure}

>>>>

\emph{The retinotectal projection, in which the axons of retinal ganglion cells form an ordered map in the optic tectum, is an important developmental system that has been studied for many decades. There is a large body of evidence, consisting of surgical and genetic manipulations, that give insight into how it behaves. Experimental work has uncovered a number of mechanisms that are involved in aspects of the mapping. However, a single, consistent theoretical framework has yet to describe all the patterns that are formed. Sperry's chemoaffinity theory provides an account of the wildtype pattern. Gierer formalised and extended this theory to incorporate competition to account for a subset of surgical manipulation experiments, but it was never tested against the full range of surgical manipulations. Other theoretical and computational studies have been partially successful in accounting for surgical manipulations, but to date, no single theory, operating only by local interactions, has accounted for every manipulation. We present a two-dimensional model of the retinotectal projection in which orthogonal pairs of opposing tectal ligand gradients combined with retinal gradients of receptors form Gierer's potential functions and then show that this receptor-ligand model accounts for the full range of surgical manipulations. We then extend the detail with which the receptor-ligand interactions are treated to successfully account for more recent genetic manipulation experiments.}

@@

Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity \citep{oleary_molecular_1999}.
In this system a two-dimensional map of  the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum. 
This retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum).
This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

@@

Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}. 
These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue. 
In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples). 
However, we found it numerically intractable to extend the model described in \citet{james_modelling_2020}, which simulated 41 whisker-specific projections, to the retinotectal system for which hundreds of projections are required (to demonstrate surgical manipulations).
Instead, we sought to implement an \emph{agent-based} model of differential chemotaxis, in which each axon growth cone is an agent that interacts with its environment and with the other growth cone agents in the system.
With this model, we asked: i) Can a two-dimensional implementation of Gierer's model explain the full range of surgical manipulations that have been described in the literature? ii) Can the model be developed such that only \emph{local} interactions are required to transfer information between agents? and iii) Can a differential chemotaxis model explain more recent results in which levels of EphA receptor sub-types are changed by genetic manipulation \citep{brown_topographic_2000,reber_relative_2004}?

@@

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in orthogonal pairs of gradients.
Although several studies model tectal ligand expression with exponential functions \citep{koulakov_stochastic_2004}, the experimental evidence for the form of tectal ligand expression is more ambiguous than that for retinal receptor expression (For example data, see Fig.\,\ref{f:expr_ephb} which shows expression patterns for EphB/ephrinB). 
We kept in mind the possibility that tectal ligand expression may be modelled by another function, but initially, we set it to the same exponential used for retinal expression and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$ (Fig.\,\ref{f:ex}).

@@

After finding an optimum set of parameters for the wildtype arrangement, we tested our model with a number of simulations of surgical manipulations (for reviews, see \citet{udin_formation_1988} and \citet{goodhill_retinotectal_1999}). 
Figs.~\ref{f:GJsurg_grafts}A and \ref{f:GJsurg_grafts}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}. 
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column). 
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression. 
The manipulated gradients, interacting with unchanged axon branches, led to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours. 
Long connecting lines in the fish net show that a number of axons from near the edges of the tectum were drawn into the graft region. 
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg_grafts}A \& B. 
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red). 
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04). 
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg_grafts}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped. 
There is considerable disorder in the simulated pattern of axon centroids but a line of red/yellow agents that would been located caudally has advanced rostrally and a number of green agents are more caudal than in the wildtype arrangement, consistent with experimental observations.

Fig.\,\ref{f:GJsurg_ablate}A shows the result of a retinal ablation, in which the temporal half of the retina is surgically removed and the remaining RGC axons are allowed to develop on the unmanipulated tectum.
According to a pure chemoaffinity theory, the surviving retinal axons should retain their chemical labels and they would be expected to find their usual locations on the tectum, resulting in a tectum which is only half populated with axons. 
The experimental observation is that the axons arrange themselves across the whole tectum, retaining their topographic order (inset, \citet{attardi_preferential_1963,schmidt_expansion_1978}). 
The fish-net plot of the simulation indicates that the competition in the system successfully `stretches out' the surviving axons to achieve a close representation of the experimentally observed result. 
In this figure, the right-hand graph shows the simulated position of each axon centroid on the rostro-caudal (R--C) tectal axis plotted against its position on the nasal-temporal (N--T) retinal axis in red triangles, with the experimental positions plotted in blue squares.
%
Fig.\,\ref{f:GJsurg_ablate}B shows the result of a tectal ablation. Here the retina is left un-modified, but the caudal half of the tectum is surgically removed. Experimental observation \citep{yoon_reorganization_1971,sharma_reformation_1972} shows that RGC axons `crush up' on the remaining rostral section, retaining topographic order and creating a complete map on the available tectal tissue. Our model clearly replicates this result, indicated by both fish-net and tectal/retinal axes graphs, because the chemotaxis component of the model is based on balanced, relative interactions between opposing gradients, rather than absolute labels or expression measurements.
%
A related ablation experiment is the so-called `mismatch', in which the temporal half of the retina is ablated along with the caudal part of the tectum to which it normally maps, again challenging the pure chemoaffinity model \citep{horder_retention_1971}. Our model replicates this manipulation (Fig.\,\ref{f:GJsurg_ablate}C) for the same reason as in the simple tectal ablation of Fig.\,\ref{f:GJsurg_ablate}B; opposing gradients remain on the surviving tectum allowing RGC axons to find their relative ordering.
%
The final surgical manipulation which we investigated was the `compound eye' experiment, in which two nasal retina halves are fused together \citep{gaze_retino-tectal_1963,fawcett_retinotectal_1982}. Within the compound retina there exist matched pairs of retinal ganglion cells, one from each nasal half, whose receptor and ligand expressions cannot be distinguished. These form overlaid maps, each of which spans the entire tectum (Fig.\,\ref{f:GJsurg_ablate}D).

@@

In the early 2000s, a series of genetic experiments added to the set of manipulations that a model of the retinotectal system would need to
explain. 
First, \citet{brown_topographic_2000} showed that a selective RGC knock-in of EphA3 receptors led to an interesting modulation of the maps found in the tectum. 
For the strongest knock-in (a homozygote manipulation) two maps were formed, with knock-in cells finding termination zones more rostral than their wildtype retinal neighbours. 
By enhancing the quantity of EphA receptors expressed by half of all retinotectal axons, they demonstrated the importance of retinal receptor expression for the arrangement of the tectal map. 
In the weaker, heterozygote manipulation, the maps were still separated, but the separation `collapsed' towards the rostral end of the tectum (see \citet{brown_topographic_2000}, Fig\,5). 
This led to a second study showing that with an additional knock-down of EphA4 receptors in \emph{all} RGCs, the retinal origin of the collapse point was shifted temporally and all heterozygote EphA3 knock-in cells were pushed even further rostrally (\citet{reber_relative_2004}, their Fig.\,3e). 
For our model to successfully reproduce these results, it should therefore: i)~Demonstrate a complete separation of maps for a `large' knock-in of EphA3; ii)~Show a separation of maps for a smaller knock-in of EphA3 that is greater for axons of nasal origin. iii)~For the smaller knock-in, a collapse point at 76\% of the N--T axis; iv)~Show a considerable rostral shift of cells for which EphA3 is knocked in and EphA4 is knocked down; and v)~Show a collapse point for the knock-in/knock-down cells and knock-down only cells at around 88\% of the N--T axis.

An assumption made in these papers is that all EphA receptor sub-types contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which increases from a low level at the nasal side of the retina, to a high level at the temporal side. 
The precise contribution of each sub-type to the overall level is assumed not to be important. 
In our model, receptor $r_0$ increases in the retina in a nasal to temporal direction (Fig.\,\ref{f:ex}B) and so we treated $r_0$ as being equivalent to $\sum\mathrm{EphA}$ and manipulated it to attempt a recreation of the \citet{brown_topographic_2000}~and \citet{reber_relative_2004}~experiments. 
In half of the RGCs, we knocked EphA3 in by adding to the value of $r_0$ in these cells either 0.8 (heterozygote EphA3 ki/+) or 3.2 (homozygote EphA3 ki/ki). 
To include the additional heterozygote knock-down of EphA4 (EphA4 +/-) from \citet{reber_relative_2004}, we subtracted 1.2 from $r_0$ in \emph{all} RGCs.
Fig.\,\ref{f:GJeph}A shows, for the single EphA3 knock-in (EphA3 ki/+), branch positions, a plot of axon centroids and final rostral-caudal position on the tectum versus the originating nasal-temporal retinal coordinate for time step 1500. 
The left panel shows branches coloured in the red/green duochrome scheme according to their origin location on the retina.
In the middle panel, axon centroids of RGCs which were treated with the knock-in manipulation are coloured red, wildtype cells are blue. 
The right hand panel mapping final R--C position versus N--T origin  (and its inset which shows the experimental result) follow the same red/blue colour scheme.
The map of treated cells (red) was shifted slightly rostrally, and the gap between the red and blue curves became smaller for more temporal RGCs. However, it can't be described as a map collapse (the term used in \citet{brown_topographic_2000}) because the curves remain separated for all N--T values. Regardless of the value of the ratio between treated and untreated cells, treated temporal axons are drawn to a significantly more rostral position than those without the knock-in. A bootstrapped studentized t-test of the two sets of points indicates a low probability, $p<10^{-4}$, that any of them were drawn from statistically identical distributions.

Fig.\,\ref{f:GJeph}B shows the same trio of graphs for the homozygote double knock-in, EphA3 ki/ki. Here, the treated axons were pushed further towards the rostral end of the tectum and the untreated axons were pushed caudally due to competition with the treated cells. 
The maps were completely separated. As in the experiment, there was no collapse point.
The rostral movement of treated axons occurred due to their increased expression of EphA ($r_0$), which increased the simulated interaction with the rostral-caudal tectal ligand gradient. 
The opposing receptor expression ($r_2$) was unchanged by the manipulation and so those axons with increased $r_0$ expression were pulled to a more rostral position than they would otherwise have taken. 
The caudal movement of wildtype axons, whose receptor expression in Figs.\,\ref{f:GJeph}A-B was unchanged, occurred due to competition with the most caudal of the rostrally-drawn treated axons.
Overall, the correspondence between the experimental result and the simulated homozygous knock-in is close.

Fig.\,\ref{f:GJeph}C shows the result for the \citet{reber_relative_2004} manipulation, in which the EphA expression was selectively enhanced (by EphA3 knock-in) in 50\% of cells and knocked down (by EphA4 knock-down) in all cells.
As in Fig.\,\ref{f:GJeph}A, the red and blue lines are separated, but there is no map collapse ($p<10^{-4}$). 
The significant rostral movement of the treated axons seen in \citet{reber_relative_2004}, where not a single treated axon was found more than 30\% along the rostral-caudal axis, was absent from this result. 
In fact, the knock-in/knock-down cells find locations slightly more \emph{caudal} than those in Fig.\,\ref{f:GJeph}A, because while $r_0$ is increased by 0.8 from the knock-in, the knock-down of 1.2 reduces these cells' interaction with tectal ephrinA ($L_0$) compared with those in the EphA3 knock-in/EphA4 wildtype simulation.

It is important to note that these results emerge purely from modifications to the receptor expression levels; the model is otherwise unchanged. 
The lack of collapse points or the more rostral positions of knock-in/knock-down cells suggests that there is an interaction effect between EphA4 and other EphA receptors that has not been captured by the differential chemotaxis with competition model.
We address possible interaction effects in the next section.

@@

$c_i\,r_{i,b}$ is the effective signal strength, $s_i$, for receptor $i$ and differs from $r_{i,b}$ only for $i=0$ for which (dropping the branch subscript) $s_0 = r_0/r_{\!\scriptscriptstyle A4}$.
Under this relationship, the signal strength changes as $r_0$ increases (simulating the knock-in of EphA3) and also as $r_{\!\scriptscriptstyle A4}$ decreases (simulating EphA4 knock-down).  
Fig.\,\ref{f:clustermech}C shows $s_0$ under knock-in, knock-down and combined knock-in/knock-down conditions.
The average magnitude of the signal $s_0$, as defined above, is smaller than $r_0$, which is the effective signal in the original model (in Eq.\,\ref{e:G}, $r_{i,b} \equiv s_i$). To counterbalance this reduction in the magnitude of the receptor-ligand interaction between $r_0$ and $l_0$, we reduced the value of $r_2$: $r_2 \rightarrow r_2/2$ which ensures that the cluster size model reproduces the wildtype result as in Fig.\,\ref{f:GJ}D.

@@

% Summarize the findings and place them in context
Can the assumptions implicit in Sperry's chemoaffinity theory account for the behaviour of the axons that grow from the retina into the optic tectum?
If not, does the addition of competition resolve the remaining problems? 
Can chemotaxis and competition be implemented by invoking a signalling system that behaves like the known signalling molecules (Ephs/ephrins, Ryk/Wnt3 \& Neogenin/RGM)?
To answer these questions, we developed a model combining a gradient-following form of chemotaxis (Eq.\,\ref{e:G}) together with axon-axon competition (Eq.\,\ref{e:X}) in which information between axonal growth cone agents was only transferred by the local interactions between their receptors and ligands.
The model contained parameters to i) control the balance of competition and chemotaxis, ii) set the interaction range of the competition and iii) set a competitive interaction threshold. 
These parameters were optimized so that the model would reproduce the experimentally observed topographic patterning of growth cones seen in healthy animals. 
The model then correctly recreated a broad range of patterns that occur when healthy animals are subjected to surgical manipulations to disrupt neuronal development. 
Qualitative rotation was observed in graft rotation simulations; the patch swap simulation was similarly qualitatively accurate. 
Retinal and tectal ablation simulations produced highly convincing results, as did mismatch and compound eye simulations.

@@

@@

@@

% FIGURE S1
\begin{figure}
\includegraphics[width=\linewidth]{./images/Tissuevisb.png}
\caption{(Supplementary) Visualisation of ligand expression and gradients on the tectum for the
graft-rotate 90 degrees simulated surgical manipulations.}
\label{f:trot90}
\end{figure}
